---
title: "R_Studio_Tutorial"
output:
  html_notebook: default
  pdf_document: default
  html_document:
    df_print: paged
---
# R Basics

## Introduction

To enter code use command+enter

### R Calculator

Basic calculations:
```{r}
1+100
```

```{r}
2^2
```

```{r}
7/2
```
```{r}
10-2
```

```{r}
2*2
```

Order of operations matter: PEMDAS
```{r}
3+5*2
(3+5)*2
```

Using scientific notation:
```{r}
2e2
2/10000
```

Can also use trigometric functions:
```{r}
log(1)
exp(0)
```


You can also ask questions: T/F Statements
```{r}
1 == 1

1 >=1

1 < 1

1 != 2
```

You can define objects/variables:
```{r}
x <- 10
x
x+x
```

```{r}
y <- x+x
y
```

### Vectorization

Use the colon *:* to make a list:
```{r}
1:5
```

Conducting operations on vectors:
```{r}
2^(1:5)
```

Storing vector as an object:
```{r}
v <- 1:5
2^v
```

Log transformations:
```{r}
log10(v)
```

### Environment

To see what is in current environment us *ls()* command:
```{r}
ls()
```

To remove items from environment, use *rm()* command:
```{r}
rm(x)
ls()
```

## Package Management

To see what is already present, use command:
```{r}
installed.packages()
```
```{r}
installed.packages("vegan", dependencies=TRUE)
```

## Project Management

1. Set up management structure (i.e., "data", "src [source]", "results", "doc"). Basic folders you should have when doing data analysis.
2. Avoiding redundancy of files
3. Make sure to use version control (i.e., Git)

## Getting Help
If you don't know what a command does, just place a question mark (?) infront of the command:
```{r}
?plot
```
You can also use the help command:
```{r}
help()
```
You can even use it with symbols:
```{r}
#? <-
```

```{r}
?"<-"
```
Can get vignette:
```{r}
vignette("FAQ-vegan")
```
To get citations:
```{r}
citation("vegan")
```
## Data and Their Formats

data.frame makes a spreadsheet (like in python)
```{r}
cats <- data.frame(coat = c("calico", "black", "tabby"), weight = c(2.1,5,3.2), likes_string = c(1,0,1))
```

```{r}
cats
```
```{r}
getwd()
```


To create a new column, use cbind:
```{r}
logweight <- log(cats$weight)
cbind(cats,logweight)
```
To print a string of text, use "paste" commmand:
```{r}
paste("My cat is ", cats$coat, ", and it weighs ", cats$weight, " kg.", sep = "")
```

### Data Types

```{r}
typeof(cats$coat)
```
```{r}
typeof(cats$weight)
```
 A  double represent a numver with decimal places
 
```{r}
typeof(cats$likes_string)
```
 
```{r}
typeof(3.14159)
```

```{r}
typeof(1i)
```

You can use "TRUE" & "FALSE"
```{r}
typeof(TRUE)
typeof(T)
typeof(FALSE)
typeof(F)
```

```{r}
typeof(cats)
```

```{r}
class(cats)
```

To open a file use:
```{r}
file.show()
```

To make 1=TRUE & 0=FALSE
```{r}
cats$likes_string <- as.logical(cats$likes_string)
typeof(cats$likes_string)
```
```{r}
cats$likes_string
```

```{r}
c("a", "b")
ab <-c("a","b")
ab
```

```{r}
c(ab, "c")
```

```{r}
c(ab, ab)
```
To get a range, just use a colon (:)
```{r}
1:10
```

You can also use seq(#):
```{r}
seq(10)
```

```{r}
z <- seq(10)
head(z, n=3)
```
```{r}
length(z)
class(z)
typeof(z)
```
```{r}
seq(1,100, by=5)
```

Gives you 1 to 100 in intervals of 5

```{r}
seq(70,100, by=5)
```

### Data Frames

```{r}
cats <- cbind(cats,logweight)
cats
```

```{r}
str(cats$coat)
```

```{r}
str(cats$coat)
```

```{r}
coats <- "tabby", "tortoiseshell" "tortoiseshell", "black", "tabby"
coats
```

```{r}
factor("coats")
```

```{r}
categories <- factor(coats)
class(coats)
class(categories)
```

### Lists

```{r}
list_example <- list(title="Numbers", numbers=1:10, data=TRUE)
list_example
```
```{r}
another_list <- list(1,"a", TRUE, 1+1i)
another_list
```
```{r}
typeof(list_example)
class(list_example)

typeof(another_list)
class(another_list)

typeof(cats)
class(cats)
```

To turrn lists -> data.frame
```{r}
data.frame(list_example)
data.frame(another_list)
```
To subset lists, # after comma means look at colomuns, # before means look at row:
```{r}
cats
```

```{r}
cats[,3]
cats[3,]
```

```{r}
cats[2:3,c(1,3)]
```

### Matrices

Matrix: made for math, all numbers

Data.frame/spreadsheet: can be quantitative or qualitative

```{r}
matrix_example <- matrix(0,ncol=5, nrow=3)
matrix_example
```

```{r}
class(matrix_example)
```

```{r}
typeof(matrix_example)
```

```{r}
str(matrix_example)
```

```{r}
dim(matrix_example)
```

```{r}
ncol(matrix_example)
```

```{r}
nrow(matrix_example)
```

To turn a matrix into a spreadsheet
```{r}
class(data.frame(matrix_example))
```

To save as a data.frame
```{r}
df_example <- data.frame(matrix_example)
df_example
```

### Subsetting
Works with dataframes & matrices

```{r}
p <- c(2.3, 6.9, 4.0, 23, 1)
p
```

TO apply names to vector use:
```{r}
names(p) <- c('a', 'b', 'c', 'd', 'e')
p
```

```{r}
p[1]
```

```{r}
p[2:4]
```

```{r}
p[c(1,5)]
```

```{r}
p[c(1,1,1,3,4,5)]
```

To remove columns, use a (-)
```{r}
p[-3]
```

```{r}
p[c(-1,-5)]
```

```{r}
p[-(2:4)]
```

```{r}
p[c('a','c')]
```

To say not equal to, use !=
```{r}
p[names(p) !='c']
```

### Factors

```{r}
f <- factor(c('a', 'b', 'c','d', 'e'))
f
```

To do absolutely is equal to, use double ='s
```{r}
f[f == 'a']
```

```{r}
f[1:3]
```

To see if the factor contains some option, use %in%
```{r}
f[f %in% c('b','c')]
```

```{r}
f[-3]
```

```{r}
f2 <- factor(c('a','a','b','c','c'))
f2
```

```{r}
f2[f2 == 'a']
```

```{r}
f2 [f2 %in% c('a','c')]
```

### Matrices Resumed
Setting the seed will make sure you get the same result everytime.
```{r}
set.seed(65)
m <- matrix(rnorm(6*4), ncol=4, nrow=6)
m
```

## Slicing Matrices
For specific rows
```{r}
m[3:4, c(3,1)]
```

For specific columns
```{r}
m[,c(3,4)]
```

### Lists

```{r}
xlist <- list(a="BIOL48006220", b=seq(1,10, by = 0.5), data="Grades")
xlist
```

by = 0.5 is step size/increment

```{r}
xlist[1:2]
```


```{r}
xlist['a']
```
You can also subslice by using [[]] to represent just the value not the object
```{r}
xlist[['a']]
```

```{r}
xlist['b']
```

```{r}
xlist[['b']]
```

You can use a $ to represent what ever the values are:
```{r}
xlist$b
```

### Date Frames
```{r}
gp <- read.csv('/Users/maggieshostak/Desktop/Practical\ Computing/Class\ Examples/PracticalComputing_2022/data/forest_area_sq_km.csv')
head(gp, n=10L)
```
```{r}
head(gp[3], n=10L)
```
To determine how many rows
```{r}
nrow(gp)
```

```{r}
head(gp["country"])
```

```{r}
head(gp[["country"]], 10L)
```

```{r}
gp$X1997
```

To just look at rows 1-3 & columns 2-5
```{r}
gp[1:3, 2:5]
```
To just look at columns 1 & 3 & 4 & 5
```{r}
gp[c(1,3), c(2,5)]
```
```{r}
gp[ which(gp$X1989 <= 100 & gp$X2019 >= 100),]
```

### Conditionals & Flow
To indent we use { then press enter to make conditional statement:
```{r}
n <- 10

if (n<10) {
  print("n is less than 10")
} else if (n>10) {
  print("n is more than 10")
} else {
  print("n is equal to 10")
}
```

```{r}
for (n in seq(1,20)) {
  if (n<10) {
    print("n is less than 10")
  } else if (n>10) {
    print("n is more than 10")
  } else {
    print("n is equal to 10")
  }
}
```

Paste is like the cat command:
```{r}
g <- 0

while (g <=10) {
  print(paste(g, "is less than or equal to 10"))
  g <- g+1
}
```

### Plotting & Figures

```{r}
#installed.packages('ggplot2')
```

```{r}
library(ggplot2)
```

```{r}
gp <- read.csv('/Users/maggieshostak/Desktop/Practical\ Computing/Class\ Examples/PracticalComputing_2022/data/gapminder_all.csv')

head(gp)
```
```{r}
ggplot(data=gp, mapping=aes(x=gdpPercap_1952, y=pop_1952)) + geom_point()
```

```{r}
ggplot(data=gp, mapping=aes(x=gdpPercap_2002, y=pop_2002)) + geom_point()
```

```{r}
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv")
head(gapminder)
```
```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country, color=continent)) + geom_line()
```


```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country)) + geom_line(mapping=aes(color=continent)) + geom_point()
```

To transfer an exponential model, use the log10:
```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point() + scale_x_log10()
```
```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(alpha=0.5) + scale_x_log10()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(alpha=0.1) + scale_x_log10()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(alpha=0.25, color="purple", size=0.9, pch=15) + scale_x_log10() + geom_smooth(method=lm, color="gold", size=1.5)
```


```{r}
Africas <- gapminder[gapminder$continent == "Africa", ]
head(Africas)

```

Facet_wrap: tells you how you want to plot them independetly 
```{r}
ggplot(data=Africas, mapping=aes(x=year, y=lifeExp)) +
  geom_line() +
  facet_wrap(~ country)
```

```{r}
ggplot(data=Africas, mapping=aes(x=year, y=lifeExp)) +
  geom_line() +
  facet_wrap(~ country) +
  theme(axis.text.x = element_text(angle=90))
```

```{r}
ggplot(data=Africas, mapping=aes(x=year, y=lifeExp)) +
  geom_line(color="red") +
  facet_wrap(~ country) +
  theme(axis.text.x = element_text(angle=90)) +
  labs(
    x="Year",
    Y="lifeExp",
    Title="Life Exoectency Over Time in African Countries",
  )
```

How to save a graph to make an output file:
```{r}
AfricanLifeExpectancy <- ggplot(data=Africas, mapping=aes(x=year, y=lifeExp)) +
  geom_line(color="red") +
  facet_wrap(~ country) +
  theme(axis.text.x = element_text(angle=90)) +
  labs(
    x="Year",
    Y="lifeExp",
    Title="Life Expectency Over Time in African Countries",
  )
```

```{r}
ggsave(filename="data/AfricanLifeExpectancy.png", plot=AfricanLifeExpectancy, width=24, height=40, dpi=300, units="cm")
```


```{r}
pdf(file="results/AfricanLifeExp.pdf", width=24, height=40)
plot(AfricanLifeExpectancy)
dev.off()
```


```{r}
write.table(gapminder, file="data/gapminder_web.csv", sep=",")
```

```{r}
write.csv(Africas, file="data/gapmminder_web_africas.csv")
```

### Fancy Plots

```{r}
#install.packages(c("ggridges", "viridis", "hrbrthemes", dependencies=T))
```
```{r}
library(ggridges)
library(ggplot2)
library(viridis)
library(hrbrthemes)
```
```{r}
hrbrthemes::import_roboto_condensed()
```

..x.. makes it a gradient fill & smooth

```{r}
ggplot(lincoln_weather, aes(x=`Mean Temperature[F]`, y= `Month`, fill=..x..)) + geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) + scale_fill_viridis(name="temp. [F]", option = "A") + theme_ipsum() +
  theme(
    legend.position="none",
    panel.spacing = unit(0.1, "lines"),
    strip.text.x = element_text(size=8))
```


Themes dictate fonts

Taking family and year described, use bin to make a hard/bar plot


```{r}
install.packages(c("ggstatsplot", "palmerpenguins", dependencies=T))
```

```{r}
library(tidyverse)
library(ggstatsplot)
```

```{r}
data("penguins", package = "palmerpenguins")
penguins
```

```{r}
penguins2 <-  drop_na(penguins)
penguins2
```
```{r}
plt <- ggbetweenstats(
  data = penguins,
  x=species,
  y=bill length)
plt
```

```{r}
gapminder2 <- drop_na(gapminder)
```

```{r}
boxplot(gapminder2$lifeExp ~ gapminder2$continent)
```
Box plots don't show distribution, instead use a violin plot
```{r}
install.packages("vioplot", dependencies = T)
```
```{r}
library(vioplot)
```

```{r}
with(gapminder2, vioplot(
  lifeExp~continent, col="orange"
))
```

```{r}
plt <- ggbetweenstats()
```

To save/export graph
```{r}
ggsave(
  flename = "insert name",
  plot = plt
  width  = 8
  height = 8
  device = "png"
)
```
 You can then load the stuff in InkScape to adjust the visual aesthetics

## PCA Plots
```{r}
penguins2
```
For principle components, you need a matrix
```{r}
penguins_matrix <- with(penguins2, cbind(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g))
penguins_matrix
```

```{r}
penguin_pca <- princomp(penguins_matrix, cor = TRUE)
summary(penguin_pca)
```
Component 1 represents # of dimensions you have (x-axis, y-axis, z-axis & 4th-axes) and it rotates/interactions it based on categories [this will change the axes]

Individual Loadings
```{r}
loadings(penguin_pca)
```

```{r}
biplot(penguin_pca, xlab=penguins2[,2])
```


```{r}
levels(penguins2$species)
```

To get PCA scores
```{r}
penguin_pca$scores
```

*ADD LESSON STUFF FROM LAST CLASS*

## Statistics

### Additional Ways of Importing, Reading, and Manipulating Data


Hard method:
```{r}
rand <- c(12, 54, 98, 65, 38)
rand
```

To get cumulative sum:
```{r}
sum(rand)
```

```{r}
length(rand)
```

To get the mean of your data:
```{r}
avg.rand <- (sum(rand)/length(rand))
avg.rand
```
Another way to get the average:

```{r}
mean(rand)
```

To sort:
```{r}
sort(rand)
```

```{r}
min(rand)
```

```{r}
max(rand)
```

```{r}
cumsum(rand)
```

To get differences:
```{r}
diff(rand)
```

Slicing/Indexing #'s out:
```{r}
rand[2]
```

Transforming #'s: adding, multiplying, etc

Importing Data, Most Useful: Take a list of numbers, don't need to add comma's and just paste in data. Press 'enter' twice to finish adding data
```{r}
peds <- scan()
peds
```

### Analyses

#### Univariate Statistics

##### 1. Categotical Data
######  a. Barplots

```{r}
beer <- c(3, 4, 1, 1, 3, 4, 3, 3, 1, 3, 2, 1, 2, 1, 2, 3, 2, 3, 1, 1, 1, 1, 4, 3, 1)
length(beer)
```
You cannot use "barplot(beer)" command right away. It will just show you every # in its own column

```{r}
table(beer)
```

```{r}
barplot(table(beer))
```
This is just for counts/frequency:
```{r}
barplot(table(beer), xlab="Beer", ylab="Frequency")
```
For proportion:
```{r}
barplot(table(beer)/length(beer), xlab="Beer", ylab="Proportion")
```
###### b. Pie Charts
```{r}
pie(table(beer), main="Beer Preference By Students")
```


##### 2. Numerical Data

###### a. Stem-and-Leaf Plots
```{r}
stem(peds)
```

###### b. Strip Chart
```{r}
stripchart(peds, method="stack")
```

##### 3. Measures of Center

###### a. Mean
```{r}
mean(peds)
```

Using Columns in Data frame:
```{r}
mean(gapminder2$lifeExp)
```

###### b. Median
```{r}
median(peds)
```

```{r}
median(gapminder2$lifeExp)
```

Must know exact column headers using the "with" command
```{r}
with(gapminder2, median(lifeExp))
```

###### c. Mode

Frequency of measurement occurring:
```{r}
which(table(peds) == max(table(peds)))
```

##### 4. Variation/Spread

###### a. Range
```{r}
range(peds)
```

Will give you a summary of the range:
```{r}
diff(range(peds))
```

###### b. Variance
```{r}
var(peds)
```

Standard Deviation: sqrt of variance
```{r}
sd(peds)
```

###### c. Inter-quartile Range (IQR)
```{r}
IQR(peds)
```

###### d. Z-Scores
The number of standard deviations away from the mean
```{r}
scale(peds)
```

###### e. Summaries
```{r}
summary(peds)
```

##### 5. Plotting

###### a. Histogram
```{r}
hist(peds)
```
Modify # of breaks:
```{r}
hist(peds, breaks=2)
```


```{r}
hist(peds, breaks="scott")
```

```{r}
hist(peds, breaks="scott", prob=TRUE) #To look at the densities
```

```{r}
hist(peds, breaks="scott", prob=TRUE)
lines(density(peds))
```

```{r}
plot(density(peds))
```

##### b. Boxplots

```{r}
boxplot(peds)
```

#### Bivariate & Multivariate Statistics

##### 1. Plotting & Regression

###### a. Boxplotting
```{r}
spid.gen <- read.csv("/Users/maggieshostak/Downloads/spider_genitalia.csv", header = TRUE)
spid.gen
```
```{r}
boxplot(spid.gen) # This is a spreadsheet so it won't work
```
```{r}
boxplot(spid.gen$left.bulb ~ spid.gen$habitat)
```
```{r}
boxplot(spid.gen$left.bulb/spid.gen$carapace.length ~ spid.gen$habitat)
```

###### b. Scatterplots

```{r}
plot(spid.gen$left.bulb ~ spid.gen$right.bulb)
```

Changing plot characters using pch
```{r}
plot(spid.gen$left.bulb ~ spid.gen$right.bulb, pch = as.character(spid.gen$habitat))
```

###### c. Linear Regression

To use a linear model, define general regression variable using lm():
```{r}
gen.reg <- lm(spid.gen$left.bulb ~ spid.gen$right.bulb)
gen.reg
```

Using summary(gen.reg) will result in an ANOVA test:
```{r}
summary(gen.reg)
```

Using abline will give you your linear regression line into your plot:
```{r}
plot(spid.gen$left.bulb ~ spid.gen$right.bulb, pch = as.character(spid.gen$habitat))
abline(gen.reg)
```

###### d. Correlation Coefficient & Spearman Rank Correlation

Cor.gen will give you your R-value, to get R^2, use cor.gen^2:
```{r}
cor.gen <- with(spid.gen, cor(left.bulb, right.bulb))
cor.gen
cor.gen^2
```

```{r}
spearman.cor.gen <- with(spid.gen, cor(left.bulb, right.bulb, method="spearman"))
spearman.cor.gen^2
```

###### e. Residuals

Residuals are the difference between a value and the line of best fit (LOBF)
```{r}
residuals(gen.reg)
```

###### f. Transformations

```{r}
plot(spid.gen$left.bulb^2~spid.gen$right.bulb^2)
```

```{r}
plot(log10(spid.gen$left.bulb)~spid.gen$right.bulb)
```
```{r}
log.left.bulb <- log10(spid.gen$left.bulb)
```


##### 2. Comparing Descrete Treatment Effects

###### a. Chi-Squared Test

```{r}
obs <- c(4,15,6,15,18,2)
exp <- c(0.16,0.17,0.16,0.17,0.17,0.17)
chisq.test(obs, p=exp)
```

```{r}
obs_weighted <- c(4,15,6,15,18,2)
obs_fair <- c(10,10,10,10,10,10)
exp <- c(0.16,0.17,0.16,0.17,0.17,0.17)
```

This will be significant:
```{r}
chisq.test(obs_weighted, p=exp)
```

This is not significant since p=value is 1:
```{r}
chisq.test(obs_fair, p=exp)
```

###### b. T-test

Null hypothesis (H0):  true mean is equal to zero, but you can set true mean to other values with "mu="

1 Sample T-Test: Student T-Test (2-Tail Test)
```{r}
t.test(spid.gen$carapace.length)
```

```{r}
mean(spid.gen$carapace.length)
```

```{r}
t.test(spid.gen$carapace.length, mu=29)
```

2 Sample T-Test:
```{r}
habitat.t2 <- t.test(spid.gen$carapace.length~spid.gen$habitat)
habitat.t2
```
```{r}
habitat.t2 <- t.test(spid.gen$left.bulb~spid.gen$habitat)
habitat.t2
```

1 Sample Test: Is one less than the other? (1-Tailed)
```{r}
habitat.tless <- t.test(spid.gen$left.bulb, alternative = "less")
habitat.tless
```

Conclusion: B is not less than A

1 Sample Test: Is one greater than the other? (1-Tailed)
```{r}
habitat.tgreater <- t.test(spid.gen$left.bulb, alternative = "greater")
habitat.tgreater
```


###### c. Analysis Of Variance (ANOVA)


```{r}
gen.lm <- lm(spid.gen$left.bulb~spid.gen$habitat)
gen.lm
```

```{r}
summary(gen.lm)
```

```{r}
anova(gen.lm)
```

```{r}
gen.anova <- aov(spid.gen$left.bulb~spid.gen$habitat)
gen.anova
```

```{r}
summary(gen.anova)
```

Note: You get the same output

```{r}
gapminder2.lifeexp.continent.anova <- aov(gapminder2$lifeExp~gapminder2$continent)
summary(gapminder2.lifeexp.continent.anova)
```

Performing a Post Hoc Test: Tukey Test
```{r}
gapminder2.tukey.two.way <- TukeyHSD(gapminder2.lifeexp.continent.anova)
gapminder2.tukey.two.way
```

###### d. Analysis of Covariance (ANCOVA)

Supplying to model other factors that might explain the differences
```{r}
boxplot(spid.gen$left.bulb~spid.gen$habitat)
```
Use a covariate by adding *factor_you_want_to_control
```{r}
gen.ancova <- lm(spid.gen$left.bulb~spid.gen$habitat*spid.gen$carapace.length)
summary(gen.ancova)
```

```{r}
gapminder2.lifeexp.continent_GDP.ancova <- aov(gapminder2$lifeExp~gapminder2$continent+gapminder2$gdpPercap)
summary(gapminder2.lifeexp.continent_GDP.ancova)
```

```{r}
install.packages("multcomp")
```
```{r}
library(multcomp)
```

```{r}
#posthoc.gapminder2.ancova <- glht(gapminder2.lifeexp.continent_GDP.ancova, linfct=mcp(technique="Tukey")) (isn't working)
```

##### 3. Principle Component Analysis

```{r}
sp.matrix <- with(spid.gen, cbind(left.bulb, right.bulb, carapace.length, leg4.length))
sp.matrix
```
```{r}
sp.pca <- princomp(sp.matrix, cor=TRUE)
summary(sp.pca)
```

```{r}
loadings(sp.pca)
```
```{r}
biplot(sp.pca)
```
```{r}
install.packages("ggfortify", dependencies = TRUE)
```

```{r}
sp.pca$scores
```


```{r}
library(ggfortify)
```

```{r}
sp.pca.plot <- autoplot(sp.pca,
                        data=spid.gen,
                        colour="habitat")
sp.pca.plot
```
# Plot Cheatsheet
```{r}
plot(1:25, rep(0.25,25), pch=1:25, col=1:25, ylim=c(0,6), cex=2, ylab="Line types (lty) 1 to 6", xlab="Plotting character (pch) 1 to 25 and colours (col) 1 to 8", main="Line types (lty), plotting characters (pch),\nand colors (col) for plot and xyplot", lab=c(25,7,2))

points(1:8, rep(0.5,8), pch=20, col=1:8, cex=4)

abline(h=1:6, lty=1:6, lwd=5, col=1:8)
```

# Looping and Conditionals

## A. If-Then Statements
```{r}
x <- 5
if(x > 0){
  print("Positive #")}
```

```{r}
x <- -5
if(x > 0){
  print("Positive #")
} else if (x < 0) {
  print("Negative #")
} else
  print("Zero")
```

%% Represents division, and == means no remainder "if its true", "if its odd"

```{r}
a <- c(5,7,2,9)
ifelse(a %% 2 == 0, "even", "odd")
```

## B. For Loops

For each of these values in Z, divide them by 2, if they don't have a remainder, it will tell you how many even numbers
```{r}
z <- c(2,5,3,9,8,11,6)
count <- 0

for(val in z) {
  if(val %% 2 == 0) count = count + 1
}
  print(count)
```

## C. While Statement

```{r}
i <- 1
while (i < 5) {
  print(i)
  i = i +1
}
```

## D. Interruptions

Breaks the loop
```{r}
x <- 1:5
for (val in x){
  if (val == 3){
    break
  }
  print(val)
}
```

If print(val) is placed before the break, it will print the value 3, then break the loop
```{r}
x <- 1:5
for (val in x){
  if (val == 3){
    print(val)
    break
  }
  #print(val)
}
```

```{r}
x <- 1:5
for (val in x){
  print(val)
  if (val == 3){
    #print(val)
    break
  }
  #print(val)
}
```

If the value is equal to 3, it will skip that value and move onto the next one
```{r}
for (val in x){
  if (val == 3) {
    next
  }
  print(val)
}
```

```{r}
x <- 1
repeat{
  print(x)
  x = x+1
  if (x == 20) {
    break
  }
}
```
# Randomization & Dataset Management (short)

## A. Dataset Management

```{r}
matrix_B <- matrix(
  c(2,4,3,1,5,7),
  nrow=3,
  ncol=2
)

matrix_B
```

Transforming Data
```{r}
t(matrix_B)
```

```{r}
matrix_C <- matrix(
  c(7,4,2),
  nrow=3,
  ncol=1
)

matrix_C
```
Putting the Matrices/Data.frame together
```{r}
matrix_BC <- cbind(matrix_B, matrix_C)
matrix_BC
```

Deconstructing the Matrices
```{r}
c(matrix_BC)
```

## B. Randomizations
Each tiime you run the script, you will get different numbers

rnorm(# samples, average, variance as standard deviation)
sample(what you want to sample, size of sample, replace = FALSE, prob = NULL)

```{r}
ndist <- rnorm(100, 50, 20)
ndist 
```
```{r}
hist(ndist)
```


```{r}
sample(ndist, 10, replace = FALSE, prob = NULL)
```

### Sampling With & Without Replacement
```{r}
test <- 1:100
test
```

No repeats:
```{r}
sort(sample(test, 50, replace = FALSE, prob = NULL))
```

Now you can have repeats:
```{r}
sort(sample(test, 50, replace = TRUE, prob = NULL))
```

